Statistical Machine Learning Models on A Bank Marketing dataset

### 2025-12-09 Slides.Rmd
Spaces
summary slides

## LogReg: Feature Selection & Linearity  #MUSSA
* Initial feature selection was performed using **Backward Stepwise Elimination** on the full model to find the most useful set of main effects.
* Our selected formula resutled:

$y \sim job + marital + education + balance + housing + $

$loan + contact + day + month + duration + $

$campaign + previous+ poutcome$
* **Residual Plots** were used to assess the linearity assumption for numeric predictors (`age`, `balance`). This led to testing a model with non-linear terms.

@@

## LogReg: Models with non-linear components and interactions
* A **quadratic term** $I(\text{balance}^4)$—was added to the formula to reduce the model's bias.
* We then tested for crucial **interaction terms** (between campaign features and demographics). 
* Our model had a cross-validated test error rate 0.09846

## LogReg: Regularization & Tuning   #MUSSA
* We explored Ridge and LASSO regularization as an alternative approach for feature selection.
* **5-Fold Cross-Validation** was used to select the optimal penalty parameter, $\lambda$, which minimized the mean CV error for each regularized model.

```{r, echo=FALSE, results='asis', fig.width=4, fig.height=3}

plot(cv.ridge) 
plot(cv.lasso)
```

## LogReg: Evaluation & Coefficients  #MUSSA
* The final selected LogReg model was evaluated on the 5-fold CV error rate, along with metrics focused on the confusion matrix.
* Models performed relatively similar. We would prefer a model with higher **sensitivity** (few false negatives)

@@

## Boosting 2/3: Hyperparameter Tuning 
- We tuned the number of trees using OOB error estimation to prevent overfitting.
- The OOB method selected **`r best_trees`** trees as optimal.
- This choice balances model complexity and generalization.
---

@@

## Discussion
* Consideration: regulatory/ethical implications in banking and the need to justify predictions (further supporting the value of interpretability).
* Advanced train/test splitting methods like **time-series analysis** were not used.
* Suggest future improvements: Exploring advanced techniques like **resampling methods** (SMOTE) to handle imbalance better or trying.

>>>>

# 1. Logistic Regression Model

## Feature Selection & Linearity  
* Initial feature selection was performed using **Backward Stepwise Elimination** on the full model to find the most useful set of main effects.
* Our selected formula resulted:

$y \sim job + marital + education + balance + housing + loan + contact +$

$day + month + duration + campaign + previous+ poutcome$

* Residual Plots were used to assess the linearity assumption for numeric predictors (`age`, `balance`). This led to testing a model with non-linear terms.

@@

## Models with non-linear components and interactions
* A **quadratic term** $I(\text{balance}^4)$—was added to the formula to reduce the model's bias.
* We then tested for crucial **interaction terms** (between campaign features and demographics). 
* Our model had a cross-validated test error rate `r round(round(cv_error_rate, 5), 4)``

## Regularization & Tuning  
* We explored Ridge and LASSO regularization as an alternative approach for feature selection.
* **5-Fold Cross-Validation** was used to select the optimal penalty parameter, $\lambda$, which minimized the mean CV error for each regularized model.

```{r, echo=FALSE, results='asis', fig.width=5, fig.height=4}
plot(cv.ridge) 
plot(cv.lasso)
```

## Evaluation 
* The final selected LogReg model was evaluated on the 5-fold CV error rate, along with metrics focused on the confusion matrix.
* Models performed relatively similar. We would prefer a model with higher *sensitivity* (few false negatives)

@@

## Boosting 2/3: Hyperparameter Tuning 
- We tuned the number of trees using OOB error estimation to prevent overfitting.
- The OOB method selected **`r best_trees`** trees as optimal.
- This choice balances model complexity and generalization.

@@

## Discussion
* Consideration: regulatory/ethical implications in banking and the need to justify predictions (further supporting the value of interpretability).
* Advanced train/test splitting methods like *time-series analysis* were not used.
* Suggest future improvements: Exploring advanced techniques like *resampling methods* (SMOTE) to handle imbalance better or trying.
